Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-09-26 docs/manuscript/paper.tex
extra small changes
do a check with smaller process noise

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc. The grey dashed line denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits. The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-1} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 4.28 \times 10^{-21} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.3 \times 10^{-23}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.0 \times 10^{-16}$ s$^{-3/2}$ for PSR J1939+2134 \newline

@@

and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.0 \times 10^{-16}$ s$^{-3/2}$. This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars. This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude. 

>>>>

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc. The grey dashed line denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits. The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline

@@

and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$. This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars. If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method. This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.
